---
title: "Mouse retinas: KCNV2-KO vs WT RNA-seq Analysis"
subtitle: "Revision 9 — KO2 excluded on knockout validation, SVA-adjusted"
author: "Miles Pufall"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
  echo: true
---

## Revision history

| File | Change |
|---|---|
| `sheila_rnaseq_KOvWT_3.qmd` | Original. KO2 and KO3 dropped on QC grounds (leaving 3 WT vs 3 KO of an original 3 vs 5); unshrunken log2FC; volcano at padj < 0.05 with a 20% FC cutoff. |
| `sheila_rnaseq_KOvWT_4.qmd` | All eight samples retained. Every analysis run twice, ± SVA, to test whether the adjustment was absorbing genotype signal. |
| `sheila_rnaseq_KOvWT_5.qmd` | Streamlined to the SVA-adjusted model throughout. Added the two-tier Müller glia gliosis panel and an over-representation suite (GO, KEGG, Reactome, WikiPathways) alongside GSEA; ± SVA comparison condensed to an appendix. |
| `sheila_rnaseq_KOvWT_6.qmd` | Gene-panel bar charts at padj < 0.05; osmosensitive panel restored; shrunken estimates reported plainly as `log2FC`. |
| `sheila_rnaseq_KOvWT_8.qmd` | *Kcnv2* knockout validation added as an explicit pre-modelling QC step; **KO2 excluded** because it shows no knockdown. KO3 retained. Dataset rebuilt, renormalised and SVA re-estimated on the 7 retained samples. Axis clipping to ±5 applies to the volcano panels only. |

| `sheila_rnaseq_KOvWT_9.qmd` | **This file.** Input paths validated up front via a `project_dir` setting, so a missing `quants/` fails immediately with a readable message instead of silently producing an empty sample table. |

## Design and rationale

Eight libraries were sequenced: **WT1–3 vs KO2–6** (3 WT, 5 KO). Revision 3
dropped KO2 and KO3 because they separated in the PCA; that decision is not
carried forward, since removing samples on the basis of a PCA computed from the
same expression data is circular.

**KO2 is excluded here on a different and non-circular basis:** it shows no
reduction in *Kcnv2* itself, so there is no evidence the knockout is present in
that animal. The analysis therefore runs on **3 WT vs 4 KO**. KO3 is retained —
it separated in the original PCA but its *Kcnv2* level sits with the true
knockouts.

The structure those two samples create is instead **modeled**: surrogate
variable analysis estimates it and it enters the design as a covariate. The
appendix documents why this is defensible here — SV1 carries only R² ≈ 0.12
with genotype, and the fold changes of the strongly-called genes are
essentially unchanged with or without the adjustment, so the SV is absorbing
technical structure rather than competing with the contrast.

Conventions used throughout:

- **Design:** `~ SV1 (+ SV2) + background`, contrast KO vs WT.
- **Effect sizes:** all reported `log2FC` values are apeglm-shrunken, which
  is simply the fold change this design supports; the unshrunken maximum-
  likelihood values are kept as `log2FC_MLE` for reference.
- **Significance:** `padj < 0.01`, no fold-change cutoff.
- **Volcano:** red up in KO, blue down, x-axis clipped to ±5 with off-scale
  genes counted in the subtitle rather than dropped. This clip applies to the
  volcano panels alone — every other figure is scaled to its own data.
- **Figure export:** every panel is written to `figures_pub/` as a 4 × 4 cm
  PDF, drawn at that size rather than scaled down. Legible versions appear
  inline.

## Trimming and Mapping

### Adapter trimming with TrimGalore! (v0.6+)

```
for i in *_R1_001.fastq.gz;
do
  SAMPLE=$(echo ${i} | sed "s/_R1_001\.fastq.gz//")
  trim_galore --paired -j 4 ${SAMPLE}_R1_001.fastq.gz ${SAMPLE}_R2_001.fastq.gz
done
```

### Alignment and quantification with Salmon (v1.10.2)

```
for i in *_R1_val_1.fq.gz;
do
  SAMPLE=$(echo ${i} | sed "s/_R1_val_1\.fq.gz//")
  salmon quant -i /Shared/pufall/mpufall/RNAseq/DataSets2015/3722_Dex_v_Pred/Trimmed/gen_v45_tx_index/ \
    -p 24 -l A -1 ${SAMPLE}_R1_val_1.fq.gz -2 ${SAMPLE}_R2_val_2.fq.gz \
    --validateMappings --seqBias --gcBias --posBias --numBootstraps 50 \
    -o quants/${SAMPLE}
done
```

## Setup

```{r setup, include=TRUE, message=FALSE}
bioc_pkgs <- c(
  "tximport", "DESeq2", "apeglm", "sva", "limma",
  "GenomicFeatures", "txdbmaker", "AnnotationDbi", "org.Mm.eg.db", "biomaRt",
  "clusterProfiler", "enrichplot", "DOSE", "ReactomePA", "fgsea",
  "vsn", "genefilter"
)

cran_pkgs <- c(
  "pheatmap", "RColorBrewer", "viridis", "ggrepel", "cowplot", "ggpubr",
  "patchwork", "msigdbr", "openxlsx", "scales", "rjson", "matrixStats",
  "tidyverse"
)

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager", repos = "https://cloud.r-project.org")
}

missing_cran <- setdiff(cran_pkgs, rownames(installed.packages()))
if (length(missing_cran) > 0) {
  message("Installing missing CRAN packages: ", paste(missing_cran, collapse = ", "))
  install.packages(missing_cran, repos = "https://cloud.r-project.org")
}

missing_bioc <- setdiff(bioc_pkgs, rownames(installed.packages()))
if (length(missing_bioc) > 0) {
  message("Installing missing Bioconductor packages: ", paste(missing_bioc, collapse = ", "))
  BiocManager::install(missing_bioc, update = FALSE, ask = FALSE)
}

invisible(lapply(c(bioc_pkgs, cran_pkgs), library, character.only = TRUE))

set.seed(1234)
```

### Constants and the publication figure system

```{r constants}
padj_cut  <- 0.01          # significance threshold for DE calls
panel_padj_cut <- 0.05     # threshold used to colour the gene-panel bar charts
vol_xlim  <- c(-5, 5)      # volcano x-axis clip -- volcanoes ONLY;
                           # every other panel is scaled to its own data
max_sv    <- 2             # cap on surrogate variables

col_up   <- "#D62728"      # red   = up in KO
col_down <- "#1F6FB4"      # blue  = down in KO
col_ns   <- "grey75"
sig_cols <- c(Up = col_up, Down = col_down, NS = col_ns)
geno_cols <- c(WT = "grey35", KO = col_up)

PUB_CM  <- 4
FIG_DIR <- "figures_pub"
dir.create(FIG_DIR, showWarnings = FALSE)

pdf_device <- if (isTRUE(capabilities("cairo"))) grDevices::cairo_pdf else grDevices::pdf

# Plot functions take pub = TRUE/FALSE. A figure drawn for a 15 cm screen and
# then scaled to 4 cm has hairline strokes and unreadable type, so the same
# plot is drawn twice at its intended size instead.
pub_style <- function(pub) {
  if (pub) {
    list(base = 5, pt = 0.30, stroke = 0.20, lab = 1.35,
         alpha = 0.55, bar_w = 0.65, err_w = 0.15)
  } else {
    list(base = 12, pt = 1.20, stroke = 0.45, lab = 2.90,
         alpha = 0.70, bar_w = 0.70, err_w = 0.25)
  }
}

theme_pub <- function(pub) {
  s <- pub_style(pub)
  theme_bw(base_size = s$base) +
    theme(
      panel.border      = element_rect(linewidth = s$stroke, color = "grey20"),
      panel.grid.minor  = element_blank(),
      panel.grid.major  = element_line(linewidth = s$stroke * 0.6),
      axis.ticks        = element_line(linewidth = s$stroke),
      axis.ticks.length = unit(if (pub) 1 else 2.5, "pt"),
      plot.title        = element_text(face = "bold",
                                       size = rel(if (pub) 1.0 else 1.05)),
      plot.subtitle     = element_text(size = rel(0.85), color = "grey30"),
      plot.margin       = margin(if (pub) 1 else 5, if (pub) 2 else 5,
                                 if (pub) 1 else 5, if (pub) 1 else 5),
      legend.key.size   = unit(if (pub) 4 else 12, "pt"),
      legend.margin     = margin(0, 0, 0, 0),
      legend.box.spacing = unit(if (pub) 1 else 5, "pt"),
      legend.text       = element_text(size = rel(0.8)),
      legend.title      = element_text(size = rel(0.85), face = "bold"),
      strip.text        = element_text(size = rel(0.85), face = "bold")
    )
}

save_pub <- function(plot, filename, cm = PUB_CM, cm_h = NULL) {
  ggsave(file.path(FIG_DIR, filename), plot,
         width = cm, height = if (is.null(cm_h)) cm else cm_h,
         units = "cm", device = pdf_device)
  invisible(file.path(FIG_DIR, filename))
}

# Draw legibly for the HTML and write the 4 cm version from the same function,
# so the two can never drift apart.
show_both <- function(fun, filename, ..., cm_h = NULL) {
  save_pub(fun(pub = TRUE, ...), filename, cm_h = cm_h)
  fun(pub = FALSE, ...)
}
```

## Mapping statistics

```{r sample-table}
# Point this at the folder holding quants/ and the annotation GTF. "." is the
# folder containing this .qmd, which is what Quarto uses as the working
# directory during a render -- so the default is correct as long as the .qmd
# sits beside the data.
project_dir <- "."

dir      <- file.path(project_dir, "quants")
gtf_file <- file.path(project_dir, "gencode.vM38.annotation.gtf.gz")

# ---------------------------------------------------------------------------
# Input validation.
#
# list.files() on a path that does not exist returns character(0) -- silently,
# with no error and no warning. An empty sample table then propagates: kable()
# prints an empty table without complaint, and the render dies several chunks
# later inside whichever function first cannot cope with zero rows. Every such
# failure looks like a parsing bug rather than a missing directory. Check here
# instead, and fail with something that names the actual problem.
# ---------------------------------------------------------------------------
if (!dir.exists(dir)) {
  stop("Cannot find the quants/ directory.\n",
       "  Looked for  : ", normalizePath(dir, mustWork = FALSE), "\n",
       "  Working dir : ", getwd(), "\n",
       "  That folder contains: ",
       paste(head(list.files(project_dir), 30), collapse = ", "), "\n",
       "Quarto renders with the .qmd's own folder as the working directory. ",
       "Either move this file next to quants/, or set `project_dir` above to ",
       "the folder that holds it.")
}

sample_dirs <- list.files(dir)

if (length(sample_dirs) == 0) {
  stop("quants/ exists at ", normalizePath(dir), " but contains no samples.")
}

sample_table <- sample_dirs %>%
  as_tibble_col(column_name = "names") %>%
  mutate(
    background = ifelse(substr(names, 1, 1) == "W", "WT", "KO"),
    rep        = as.integer(str_sub(names, -1)),
    background = factor(background, levels = c("WT", "KO"))
  ) %>%
  arrange(background, rep)

sample_table$files <- file.path(dir, sample_table$names, "quant.sf")
sample_table$meta  <- file.path(dir, sample_table$names, "aux_info", "meta_info.json")

# Both files must exist for every sample before anything else runs.
for (col in c("files", "meta")) {
  absent <- sample_table$names[!file.exists(sample_table[[col]])]
  if (length(absent) > 0) {
    stop("Missing ", col, " for: ", paste(absent, collapse = ", "),
         "\n  Expected e.g. ", sample_table[[col]][1])
  }
}

if (!file.exists(gtf_file)) {
  stop("Cannot find the annotation GTF.\n  Looked for : ",
       normalizePath(gtf_file, mustWork = FALSE),
       "\n  Working dir: ", getwd())
}

cat("Found ", nrow(sample_table), " samples in ", normalizePath(dir), "\n", sep = "")

knitr::kable(dplyr::select(sample_table, names, background, rep),
             caption = "Samples included in this analysis")
```

```{r mapping-stats}
# One row per sample, built explicitly.
#
# The sapply(...) + t() idiom is fragile: sapply only simplifies to a matrix
# when every meta_info.json returns a list of identical length. Salmon omits or
# adds fields depending on flags and version, so one sample with an extra field
# makes sapply return a plain list, t() yields a 1 x n matrix, and everything
# downstream silently collapses to a single row.

drop_fields <- c("quant_errors", "eq_class_properties", "length_classes")

read_meta <- function(x) {
  mi <- rjson::fromJSON(file = sample_table$meta[match(x, sample_table$names)])
  mi <- mi[setdiff(names(mi), drop_fields)]

  # Keep atomic scalars only. as_tibble() on the filtered list is not safe: if
  # the filter leaves nothing, it returns a 0 x 0 tibble, mutate() then adds a
  # column to zero rows, and map_dfr binds a frame with no `sample` column at
  # all -- at which point `sample` in the next mutate() resolves to base::sample
  # and R reports the baffling "cannot coerce type 'closure'". Building the row
  # from a guaranteed one-row tibble removes that whole failure path.
  is_scalar <- vapply(mi, function(v) is.atomic(v) && length(v) == 1, logical(1))
  mi <- mi[is_scalar]

  out <- tibble::tibble(sample = x)
  for (nm in names(mi)) out[[nm]] <- mi[[nm]]
  out
}

rs_tidy_tbl <- map_dfr(sample_table$names, read_meta)

if (!"sample" %in% names(rs_tidy_tbl)) {
  stop("No `sample` column after parsing meta_info.json. Columns found: ",
       paste(names(rs_tidy_tbl), collapse = ", "))
}

cat("Parsed fields:", paste(setdiff(names(rs_tidy_tbl), "sample"), collapse = ", "), "\n")

rs_tidy_tbl <- rs_tidy_tbl %>%
  mutate(sample = factor(sample, levels = sample_table$names))

needed <- c("num_processed", "num_mapped", "percent_mapped")
missing_fields <- setdiff(needed, names(rs_tidy_tbl))

if ("percent_mapped" %in% missing_fields &&
    all(c("num_processed", "num_mapped") %in% names(rs_tidy_tbl))) {
  rs_tidy_tbl <- rs_tidy_tbl %>%
    mutate(percent_mapped = 100 * as.numeric(num_mapped) / as.numeric(num_processed))
  missing_fields <- setdiff(missing_fields, "percent_mapped")
}

if (length(missing_fields) > 0) {
  stop("meta_info.json is missing required field(s): ",
       paste(missing_fields, collapse = ", "),
       "\nFields found: ", paste(names(rs_tidy_tbl), collapse = ", "))
}

rs_tidy_tbl <- rs_tidy_tbl %>%
  mutate(across(all_of(needed), as.numeric),
         mapped_M = num_mapped / 1e6)

if (nrow(rs_tidy_tbl) != nrow(sample_table)) {
  stop("Parsed ", nrow(rs_tidy_tbl), " mapping-stat rows for ",
       nrow(sample_table), " samples.")
}

ave_reads  <- round(mean(rs_tidy_tbl$num_processed), 0)
ave_mapped <- round(mean(rs_tidy_tbl$percent_mapped), 0)
```

Mean depth `r format(ave_reads, big.mark = ",")` read pairs; mean mapping rate
`r ave_mapped`%.

```{r plot-mapping, fig.width=10, fig.height=4}
p_mapping <- function(pub, yvar, ylab, title) {
  s <- pub_style(pub)
  ggplot(rs_tidy_tbl, aes(sample, .data[[yvar]], fill = .data[[yvar]])) +
    geom_col(color = "white", width = s$bar_w, linewidth = s$stroke) +
    scale_fill_viridis_c(option = "mako", begin = 0.3, end = 0.85, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    labs(title = title, x = NULL, y = ylab) +
    theme_pub(pub) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          panel.grid.major.x = element_blank())
}

p_reads <- show_both(p_mapping, "qc_mapped_reads.pdf",
                     yvar = "mapped_M", ylab = "Mapped reads (M)",
                     title = "Mapped reads")
p_rate  <- show_both(p_mapping, "qc_mapping_rate.pdf",
                     yvar = "percent_mapped", ylab = "Percent mapped (%)",
                     title = "Mapping rate")

p_reads | p_rate
```

## Import into DESeq2

```{r import-txi, message=TRUE}
# Paths and existence were validated in the sample-table chunk.
make_txdb_from_gff <- if (requireNamespace("txdbmaker", quietly = TRUE)) {
  txdbmaker::makeTxDbFromGFF
} else {
  GenomicFeatures::makeTxDbFromGFF
}

txdb    <- make_txdb_from_gff(gtf_file)
k       <- keys(txdb, keytype = "TXNAME")
tx2gene <- AnnotationDbi::select(txdb, k, "GENEID", "TXNAME")

tx2gene$TXNAME <- sub("\\.\\d+$", "", tx2gene$TXNAME)
tx2gene$GENEID <- sub("\\.\\d+$", "", tx2gene$GENEID)

txi <- tximport(sample_table$files, type = "salmon",
                tx2gene = tx2gene, ignoreTxVersion = TRUE)
colnames(txi$counts) <- sample_table$names
```

```{r build-dds-all}
dds_all <- DESeqDataSetFromTximport(txi,
                                    colData = as.data.frame(sample_table),
                                    design  = ~ background)
colnames(dds_all) <- sample_table$names

# Pre-filter: at least 10 counts in at least 3 samples (the smaller group size).
keep    <- rowSums(counts(dds_all) >= 10) >= 3
dds_all <- dds_all[keep, ]
dds_all <- estimateSizeFactors(dds_all)

cat("Genes retained after filtering:", nrow(dds_all), "\n")
cat("Samples:", ncol(dds_all), "\n")
```

## Sample QC

```{r qc-pca, fig.width=6, fig.height=5}
vsd_all <- vst(dds_all, blind = TRUE)

pca_plot <- function(pub, mat, coldata, title, ntop = 5000, label = TRUE) {
  s   <- pub_style(pub)
  rv  <- matrixStats::rowVars(mat)
  sel <- order(rv, decreasing = TRUE)[seq_len(min(ntop, length(rv)))]
  pca <- prcomp(t(mat[sel, ]))
  pv  <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)
  df  <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2],
                    name = colnames(mat), background = coldata$background)

  p <- ggplot(df, aes(PC1, PC2, color = background)) +
    geom_point(size = s$pt * 2.5) +
    scale_color_manual(values = geno_cols, name = NULL) +
    labs(title = title,
         x = paste0("PC1 (", pv[1], "%)"),
         y = paste0("PC2 (", pv[2], "%)")) +
    theme_pub(pub)

  if (label) {
    p <- p + ggrepel::geom_text_repel(aes(label = name), size = s$lab,
                                      box.padding = 0.3, segment.linewidth = s$stroke,
                                      segment.color = "grey60", max.overlaps = 15,
                                      show.legend = FALSE)
  }
  p
}

pca_all <- show_both(pca_plot, "qc_pca_uncorrected.pdf",
                     mat = assay(vsd_all), coldata = colData(dds_all),
                     title = "PCA: uncorrected")
pca_all
```

```{r qc-distances, fig.width=6, fig.height=5}
sampleDists      <- dist(t(assay(vsd_all)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- colnames(vsd_all)
colnames(sampleDistMatrix) <- NULL

dist_args <- function(fontsize, main) {
  list(mat = sampleDistMatrix,
       clustering_distance_rows = sampleDists,
       clustering_distance_cols = sampleDists,
       col = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
       main = main, fontsize = fontsize, border_color = NA)
}

pdf_device(file.path(FIG_DIR, "qc_sample_distances.pdf"),
           width = PUB_CM / 2.54, height = PUB_CM / 2.54)
do.call(pheatmap, dist_args(3.5, "Sample distances"))
invisible(dev.off())

do.call(pheatmap, dist_args(10, "Sample-to-sample distances"))
```

```{r qc-cooks, fig.width=6, fig.height=4.5}
# Cook's distance flags counts with outsized leverage on the fit -- a more
# specific diagnosis than "it sits apart in the PCA".
dds_cook <- estimateDispersions(dds_all, quiet = TRUE)
dds_cook <- nbinomWaldTest(dds_cook)

cooks_df <- as.data.frame(log10(assays(dds_cook)[["cooks"]])) %>%
  pivot_longer(everything(), names_to = "sample", values_to = "log10_cooks") %>%
  mutate(sample = factor(sample, levels = colnames(dds_cook)))

p_cooks_fun <- function(pub) {
  s <- pub_style(pub)
  ggplot(cooks_df, aes(sample, log10_cooks, fill = sample)) +
    geom_boxplot(outlier.size = s$pt * 0.5, outlier.alpha = 0.15,
                 linewidth = s$stroke, show.legend = FALSE) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.85) +
    labs(title = "Cook's distance", x = NULL,
         y = expression(log[10]~"Cook's D")) +
    theme_pub(pub) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid.major.x = element_blank())
}

show_both(p_cooks_fun, "qc_cooks_distance.pdf")
```

```{r qc-detection, fig.width=6, fig.height=4.5}
detected <- tibble(
  sample     = colnames(dds_all),
  detected   = colSums(counts(dds_all) >= 10),
  background = colData(dds_all)$background
) %>% mutate(sample = factor(sample, levels = colnames(dds_all)))

p_detected_fun <- function(pub) {
  s <- pub_style(pub)
  ggplot(detected, aes(sample, detected, fill = background)) +
    geom_col(width = s$bar_w, color = "white", linewidth = s$stroke) +
    scale_fill_manual(values = geno_cols, name = NULL) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                       labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
    labs(title = "Genes detected", x = NULL, y = "Genes (>=10 counts)") +
    theme_pub(pub) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid.major.x = element_blank())
}

show_both(p_detected_fun, "qc_genes_detected.pdf")
```

## Knockout validation and sample exclusion

Before any modelling, confirm the knockout is actually present in each animal
labelled KO. This is the one sample-level exclusion criterion in the analysis,
and it is deliberately **external to the expression structure**: it asks whether
the experimental manipulation worked in a given animal, not whether that animal
resembles its group-mates transcriptome-wide. That distinction is what separates
it from the PCA-based exclusion in revision 3, which used the data to decide
which data to keep.

```{r ko-validation, fig.width=6, fig.height=4.5}
ko_gene <- "Kcnv2"

ko_ens <- suppressMessages(
  AnnotationDbi::mapIds(org.Mm.eg.db, keys = ko_gene, column = "ENSEMBL",
                        keytype = "SYMBOL", multiVals = "first"))

if (is.na(ko_ens) || !ko_ens %in% rownames(dds_all)) {
  stop(ko_gene, " was not found in the filtered count matrix; knockout ",
       "validation cannot be performed.")
}

ko_df <- tibble(
  sample     = colnames(dds_all),
  background = colData(dds_all)$background,
  count      = counts(dds_all, normalized = TRUE)[ko_ens, ]
) %>% mutate(sample = factor(sample, levels = colnames(dds_all)))

# Decision rule, stated before it is applied: a KO animal whose Kcnv2 level
# reaches the lowest WT shows no evidence of knockdown. Deriving the threshold
# from the WT range rather than hard-coding a sample name keeps the rule
# reproducible and makes it obvious if a future dataset flags something else.
min_wt  <- min(ko_df$count[ko_df$background == "WT"])
ko_flag <- ko_df %>%
  dplyr::filter(background == "KO", count >= min_wt) %>%
  pull(sample) %>% as.character()

knitr::kable(
  ko_df %>%
    mutate(pct_of_WT_min = round(100 * count / min_wt, 1),
           knockdown = case_when(background == "WT" ~ "-",
                                 count < min_wt ~ "yes",
                                 TRUE ~ "NO")) %>%
    dplyr::select(sample, background, count, pct_of_WT_min, knockdown),
  digits = 0, caption = paste0(ko_gene, " normalized counts, all samples"))

p_ko_fun <- function(pub) {
  s <- pub_style(pub)
  ggplot(ko_df, aes(background, count, color = background)) +
    geom_hline(yintercept = min_wt, linetype = "dashed",
               color = "grey40", linewidth = s$stroke) +
    geom_point(position = position_jitter(width = 0.12, seed = 1),
               size = s$pt * 2.5, alpha = 0.9) +
    ggrepel::geom_text_repel(aes(label = sample), size = s$lab,
                             position = position_jitter(width = 0.12, seed = 1),
                             segment.color = "grey60", show.legend = FALSE) +
    scale_color_manual(values = geno_cols, guide = "none") +
    scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
    labs(title = ko_gene, subtitle = "dashed line = lowest WT",
         x = NULL, y = "Normalized count") +
    theme_pub(pub)
}

show_both(p_ko_fun, "ko_validation_kcnv2.pdf")

if (length(ko_flag) > 0) {
  cat("\nKO samples with NO evidence of Kcnv2 knockdown: ",
      paste(ko_flag, collapse = ", "), "\n", sep = "")
} else {
  cat("\nAll KO samples show reduced Kcnv2.\n")
}
```

```{r exclusion-statement}
excluded_samples <- ko_flag

if (length(excluded_samples) > 0) {
  cat("EXCLUDED FROM ALL DOWNSTREAM ANALYSIS: ",
      paste(excluded_samples, collapse = ", "),
      "\nReason: no reduction in ", ko_gene,
      " relative to the WT range, i.e. the knockout is not detectable in ",
      "this animal.\n", sep = "")
}

sample_table_keep <- sample_table %>%
  dplyr::filter(!names %in% excluded_samples)

cat("\nRetained: ", nrow(sample_table_keep), " samples (",
    sum(sample_table_keep$background == "WT"), " WT, ",
    sum(sample_table_keep$background == "KO"), " KO)\n", sep = "")
```

::: {.callout-important}
## Sample exclusion

**KO2 is excluded from all downstream analysis.** Its *Kcnv2* level is not
merely within the WT range but above every WT sample, which is not what an
animal carrying the knockout allele should look like. The most likely
explanations are a mis-genotyped animal or a sample swap at collection or
library preparation.

The exclusion is applied **before** normalisation, filtering, dispersion
estimation and surrogate variable analysis, because all four use every sample
in the object. Dropping KO2 only at the results stage would leave its influence
baked into the size factors and the surrogate variables.

KO3 is **retained**. It separated alongside KO2 in the original PCA, but its
*Kcnv2* level sits with the other true knockouts, so the knockout-verification
criterion keeps it. Excluding it would require falling back on the PCA argument
this re-analysis rejects.

Worth confirming independently by genotyping PCR if DNA remains — a
transcript-level result is suggestive of a swap, not proof of one.
:::

### Rebuilding the dataset on the retained samples

```{r build-dds}
# Rebuilt from the Salmon output rather than subset from dds_all, so that size
# factors, the low-count filter and every subsequent estimate are computed on
# the retained samples alone.
txi_keep <- tximport(sample_table_keep$files, type = "salmon",
                     tx2gene = tx2gene, ignoreTxVersion = TRUE)
colnames(txi_keep$counts) <- sample_table_keep$names

dds_pre <- DESeqDataSetFromTximport(txi_keep,
                                    colData = as.data.frame(sample_table_keep),
                                    design  = ~ background)
colnames(dds_pre) <- sample_table_keep$names

# Same rule as before: at least 10 counts in at least 3 samples (smaller group).
keep    <- rowSums(counts(dds_pre) >= 10) >= 3
dds_pre <- dds_pre[keep, ]
dds_pre <- estimateSizeFactors(dds_pre)

cat("Genes retained after filtering:", nrow(dds_pre), "\n")
cat("Samples in the model:", ncol(dds_pre), "\n")
```

## Surrogate variables

```{r sva}
dat <- counts(dds_pre, normalized = TRUE)
dat <- dat[rowMeans(dat) > 1, ]

mod  <- model.matrix(~ background, data = as.data.frame(colData(dds_pre)))
mod0 <- model.matrix(~ 1,          data = as.data.frame(colData(dds_pre)))

n_sv_est <- sva::num.sv(dat, mod, method = "be")
n_sv     <- min(n_sv_est, max_sv)

cat("SVs suggested by num.sv (be):", n_sv_est, "\n")
cat("SVs used:", n_sv, "\n")

if (n_sv > 0) {
  svseq  <- sva::svaseq(dat, mod, mod0, n.sv = n_sv)
  sv_mat <- svseq$sv
  colnames(sv_mat) <- paste0("SV", seq_len(n_sv))
  rownames(sv_mat) <- colnames(dds_pre)
} else {
  sv_mat <- NULL
  message("No surrogate variables detected; the model reduces to ~ background.")
}
```

An SV orthogonal to genotype can only remove noise; the more of the genotype
contrast it contains, the more real effect it removes alongside. `R²` is from
regressing each SV on genotype — near 0 is safe.

```{r sv-genotype-overlap}
if (!is.null(sv_mat)) {
  sv_geno <- map_dfr(seq_len(ncol(sv_mat)), function(i) {
    fit <- lm(sv_mat[, i] ~ colData(dds_pre)$background)
    tibble(SV       = colnames(sv_mat)[i],
           R2       = summary(fit)$r.squared,
           p_value  = anova(fit)$`Pr(>F)`[1],
           cor_geno = cor(sv_mat[, i], as.numeric(colData(dds_pre)$background)))
  })
  knitr::kable(sv_geno, digits = 3,
               caption = "Association of each surrogate variable with genotype")
}
```

```{r sva-loadings, fig.width=7, fig.height=4}
p_sv_fun <- function(pub) {
  s <- pub_style(pub)
  sv_df <- as_tibble(sv_mat) %>%
    mutate(sample = colnames(dds_pre),
           background = colData(dds_pre)$background) %>%
    pivot_longer(starts_with("SV"), names_to = "SV", values_to = "loading") %>%
    mutate(sample = factor(sample, levels = colnames(dds_pre)))

  ggplot(sv_df, aes(sample, loading, fill = background)) +
    geom_col(width = s$bar_w, color = "white", linewidth = s$stroke) +
    geom_hline(yintercept = 0, color = "grey30", linewidth = s$stroke) +
    scale_fill_manual(values = geno_cols, name = NULL) +
    facet_wrap(~ SV, ncol = 1) +
    labs(title = "SV loadings", x = NULL, y = "Loading") +
    theme_pub(pub) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid.major.x = element_blank())
}

if (!is.null(sv_mat)) show_both(p_sv_fun, "qc_sv_loadings.pdf")
```

```{r pca-corrected, fig.width=11, fig.height=5}
# A fresh VST on the retained samples -- vsd_all was computed on all eight and
# its dispersion trend still carries the excluded library.
vsd <- vst(dds_pre, blind = FALSE)

pca_keep <- show_both(pca_plot, "qc_pca_retained.pdf",
                      mat = assay(vsd), coldata = colData(dds_pre),
                      title = "PCA: retained samples")

# Visualization only -- the model uses the covariates directly, not this matrix.
if (!is.null(sv_mat)) {
  mat_corr <- limma::removeBatchEffect(
    assay(vsd), covariates = sv_mat,
    design = model.matrix(~ background, data = as.data.frame(colData(dds_pre)))
  )
  pca_corr <- show_both(pca_plot, "qc_pca_sv_corrected.pdf",
                        mat = mat_corr, coldata = colData(dds_pre),
                        title = "PCA: SV-corrected")
  pca_keep | pca_corr
} else {
  mat_corr <- assay(vsd)
  pca_keep
}
```

## Differential expression: KO vs WT

```{r deseq}
dds <- dds_pre
if (!is.null(sv_mat)) {
  for (i in seq_len(ncol(sv_mat))) {
    colData(dds)[[colnames(sv_mat)[i]]] <- sv_mat[, i]
  }
  design(dds) <- as.formula(
    paste("~", paste(c(colnames(sv_mat), "background"), collapse = " + "))
  )
}

cat("Design:", deparse(design(dds)), "\n")

dds     <- DESeq(dds)
res_mle <- results(dds, name = "background_KO_vs_WT", alpha = padj_cut)
summary(res_mle, alpha = padj_cut)

# apeglm shrinkage. Passing res = res_mle keeps the Wald p-values and the
# independent-filtering-adjusted padj, so only effect sizes change.
res_ko <- lfcShrink(dds, coef = "background_KO_vs_WT",
                    type = "apeglm", res = res_mle, quiet = TRUE)
```

### Annotation and export

```{r annotation}
# biomaRt returns "" (not NA) for genes with no symbol; an empty string is a
# legal-but-meaningless name that breaks fgsea and creates phantom rows in
# joins. Normalize blanks to NA at the source.
blank_to_na <- function(x) {
  x <- as.character(x)
  x[!nzchar(trimws(x))] <- NA_character_
  x
}

anno_map <- function(ids) {
  pull_map <- function(col) {
    suppressMessages(mapIds(org.Mm.eg.db, keys = ids, column = col,
                            keytype = "ENSEMBL", multiVals = "first"))
  }

  symbol      <- blank_to_na(pull_map("SYMBOL"))
  entrez      <- blank_to_na(pull_map("ENTREZID"))
  description <- blank_to_na(pull_map("GENENAME"))
  missing_any <- is.na(symbol) | is.na(entrez) | is.na(description)

  if (any(missing_any)) {
    ann <- tryCatch({
      ensembl <- biomaRt::useEnsembl("genes", dataset = "mmusculus_gene_ensembl")
      biomaRt::getBM(
        attributes = c("ensembl_gene_id", "external_gene_name",
                       "entrezgene_id", "description"),
        filters = "ensembl_gene_id", values = ids[missing_any], mart = ensembl)
    }, error = function(e) {
      message("biomaRt unavailable (", conditionMessage(e),
              "); keeping org.Mm.eg.db annotation only.")
      NULL
    })

    if (!is.null(ann) && nrow(ann) > 0) {
      idx  <- match(ids, ann$ensembl_gene_id)
      fill <- function(x, from) ifelse(is.na(x), blank_to_na(from)[idx], x)
      symbol      <- fill(symbol,      ann$external_gene_name)
      entrez      <- fill(entrez,      as.character(ann$entrezgene_id))
      description <- fill(description, ann$description)
    }
  }

  tibble(ensemblID   = ids,
         symbol      = blank_to_na(unname(symbol)),
         entrez      = blank_to_na(unname(entrez)),
         description = blank_to_na(unname(description)))
}

gene_anno <- anno_map(rownames(dds))

cat("Genes with a usable symbol: ", sum(!is.na(gene_anno$symbol)),
    " of ", nrow(gene_anno), "\n", sep = "")
cat("Genes with a usable Entrez ID: ", sum(!is.na(gene_anno$entrez)),
    " of ", nrow(gene_anno), "\n", sep = "")

res_tbl <- as.data.frame(res_ko) %>%
  rownames_to_column("ensemblID") %>%
  as_tibble() %>%
  dplyr::rename(log2FC = log2FoldChange) %>%
  mutate(
    log2FC_MLE = res_mle$log2FoldChange[match(ensemblID, rownames(res_mle))],
    lfcSE_MLE  = res_mle$lfcSE[match(ensemblID, rownames(res_mle))],
    stat_MLE   = res_mle$stat[match(ensemblID, rownames(res_mle))]
  ) %>%
  left_join(gene_anno, by = "ensemblID") %>%
  relocate(symbol, description, entrez, ensemblID) %>%
  arrange(padj)

sig_tbl <- res_tbl %>% dplyr::filter(padj < padj_cut)

cat("\nSignificant at padj < ", padj_cut, ": ", nrow(sig_tbl),
    " (", sum(sig_tbl$log2FC > 0), " up, ",
    sum(sig_tbl$log2FC < 0), " down)\n", sep = "")

write_csv(res_tbl, "KOvWT_SVA_results.csv")
openxlsx::write.xlsx(
  list(all_genes = res_tbl,
       sig_up    = sig_tbl %>% dplyr::filter(log2FC > 0),
       sig_down  = sig_tbl %>% dplyr::filter(log2FC < 0)),
  file = "KOvWT_SVA_results.xlsx", overwrite = TRUE)

knitr::kable(
  sig_tbl %>%
    dplyr::select(symbol, log2FC, lfcSE, padj, baseMean) %>%
    head(25),
  digits = 3, caption = "Top 25 genes by adjusted p-value")
```

### Kcnv2 in the fitted model

```{r kcnv2-result}
kcnv2_res <- res_tbl %>% dplyr::filter(symbol == ko_gene)

if (nrow(kcnv2_res) > 0) {
  knitr::kable(kcnv2_res %>%
                 dplyr::select(symbol, baseMean, log2FC, lfcSE, pvalue, padj),
               digits = 4,
               caption = paste0(ko_gene, " in the retained-sample model"))
} else {
  message(ko_gene, " did not survive count filtering in the retained samples.")
}
```

## Volcano plot

```{r volcano, fig.width=7, fig.height=6}
make_volcano <- function(pub, tbl, title, label_n = 15) {
  s  <- pub_style(pub)
  df <- tbl %>%
    dplyr::filter(!is.na(log2FC), !is.na(padj)) %>%
    mutate(cat = case_when(
      padj < padj_cut & log2FC > 0 ~ "Up",
      padj < padj_cut & log2FC < 0 ~ "Down",
      TRUE ~ "NS"))

  n_up   <- sum(df$cat == "Up")
  n_down <- sum(df$cat == "Down")
  n_off  <- sum(df$log2FC < vol_xlim[1] | df$log2FC > vol_xlim[2])

  sub <- paste0(n_up, " up, ", n_down, " down")
  if (n_off > 0) sub <- paste0(sub, "; ", n_off, " outside ±5")

  lab_df <- if (label_n > 0) {
    df %>%
      dplyr::filter(cat != "NS",
                    dplyr::between(log2FC, vol_xlim[1], vol_xlim[2])) %>%
      slice_min(padj, n = label_n, with_ties = FALSE)
  } else df[0, ]

  p <- ggplot(df, aes(log2FC, -log10(padj), color = cat)) +
    geom_point(alpha = s$alpha, size = s$pt) +
    scale_color_manual(values = sig_cols, breaks = c("Up", "NS", "Down"),
                       labels = c("Up in KO", "n.s.", "Down in KO"),
                       name = paste0("padj < ", padj_cut)) +
    geom_hline(yintercept = -log10(padj_cut), linetype = "dashed",
               color = "grey40", linewidth = s$stroke) +
    geom_vline(xintercept = 0, linetype = "dotted",
               color = "grey60", linewidth = s$stroke) +
    scale_x_continuous(breaks = seq(vol_xlim[1], vol_xlim[2], by = 2.5)) +
    coord_cartesian(xlim = vol_xlim) +
    labs(title = title, subtitle = sub,
         x = expression(log[2]~"(KO / WT)"), y = expression(-log[10]~padj)) +
    theme_pub(pub) +
    guides(color = guide_legend(override.aes = list(size = s$pt * 2.2, alpha = 1)))

  if (nrow(lab_df) > 0) {
    p <- p + ggrepel::geom_text_repel(
      data = lab_df, aes(label = symbol), size = s$lab, fontface = "italic",
      max.overlaps = 20, box.padding = 0.25, segment.color = "grey60",
      segment.linewidth = s$stroke, show.legend = FALSE)
  }

  if (pub) {
    p + theme(legend.position = "bottom", legend.direction = "horizontal")
  } else {
    p + theme(legend.position = "inside", legend.position.inside = c(0.15, 0.88),
              legend.background = element_rect(fill = alpha("white", 0.7), color = NA))
  }
}

save_pub(make_volcano(TRUE, res_tbl, "KCNV2-KO vs WT", label_n = 8), "volcano.pdf")
save_pub(make_volcano(TRUE, res_tbl, "KCNV2-KO vs WT", label_n = 0), "volcano_nolabels.pdf")

make_volcano(FALSE, res_tbl, "KCNV2-KO vs WT — SVA-adjusted")
```

## Diagnostics

```{r diagnostics, fig.width=13, fig.height=5}
p_shrink_fun <- function(pub) {
  s  <- pub_style(pub)
  df <- res_tbl %>% dplyr::filter(!is.na(log2FC), !is.na(log2FC_MLE))

  ggplot(df, aes(log2FC_MLE, log2FC, color = log10(baseMean))) +
    geom_point(alpha = s$alpha * 0.7, size = s$pt) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "grey40", linewidth = s$stroke) +
    scale_color_viridis_c(option = "viridis", name = expression(log[10]~baseMean)) +
    labs(title = "Shrinkage", x = "log2FC (MLE)", y = "log2FC (apeglm)") +
    theme_pub(pub) +
    theme(legend.position = if (pub) "none" else "right")
}

p_hist_fun <- function(pub) {
  s <- pub_style(pub)
  ggplot(as.data.frame(res_mle) %>% dplyr::filter(!is.na(pvalue)), aes(pvalue)) +
    geom_histogram(binwidth = 0.025, fill = "#3B4A6B",
                   color = "white", linewidth = s$stroke * 0.7) +
    scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
    labs(title = "p-value distribution", x = "p-value", y = "Genes") +
    theme_pub(pub)
}

p_ma_fun <- function(pub) {
  s  <- pub_style(pub)
  df <- res_tbl %>%
    dplyr::filter(baseMean > 0, !is.na(log2FC)) %>%
    mutate(dir = case_when(is.na(padj) | padj >= padj_cut ~ "NS",
                           log2FC > 0 ~ "Up", TRUE ~ "Down"))

  ggplot(df, aes(baseMean, log2FC, color = dir)) +
    geom_point(alpha = s$alpha * 0.8, size = s$pt) +
    geom_hline(yintercept = 0, color = "grey30", linewidth = s$stroke) +
    scale_x_log10(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
    scale_color_manual(values = sig_cols, breaks = c("Up", "NS", "Down"),
                       labels = c("Up", "n.s.", "Down"), name = NULL) +
    labs(title = "MA plot", x = "Mean normalized counts",
         y = expression(log[2]~"(KO / WT)")) +
    theme_pub(pub) +
    theme(legend.position = if (pub) "none" else "bottom")
}

d1 <- show_both(p_shrink_fun, "diag_shrinkage.pdf")
d2 <- show_both(p_hist_fun,   "diag_pvalue_hist.pdf")
d3 <- show_both(p_ma_fun,     "diag_ma_plot.pdf")

d1 | d2 | d3
```

```{r diagnostics-disp, fig.width=6, fig.height=5}
pdf_device(file.path(FIG_DIR, "diag_dispersion.pdf"),
           width = PUB_CM / 2.54, height = PUB_CM / 2.54, pointsize = 4)
par(mar = c(2.2, 2.4, 1.2, 0.4), mgp = c(1.3, 0.35, 0), tcl = -0.15,
    cex.axis = 0.8, cex.lab = 0.9)
plotDispEsts(dds, main = "Dispersion", cex = 0.2, legend = FALSE)
invisible(dev.off())

plotDispEsts(dds, main = "Dispersion estimates")
```

## Heatmap of top differential genes

```{r heatmap}
ann_col <- data.frame(Genotype = colData(dds)$background,
                      row.names = colnames(dds))
hm_palette <- colorRampPalette(c(col_down, "white", col_up))(255)

draw_heatmap <- function(n_genes, fontsize, main, filename = NA) {
  top <- sig_tbl %>% slice_min(padj, n = n_genes, with_ties = FALSE)
  if (nrow(top) < 2) {
    message("Fewer than two significant genes; skipping heatmap.")
    return(invisible(NULL))
  }

  mat <- mat_corr[top$ensemblID, , drop = FALSE]
  mat <- t(scale(t(mat)))
  rownames(mat) <- ifelse(is.na(top$symbol), top$ensemblID, top$symbol)

  args <- list(mat = mat, annotation_col = ann_col,
               annotation_colors = list(Genotype = geno_cols),
               color = hm_palette, show_rownames = TRUE,
               fontsize = fontsize, fontsize_row = fontsize,
               fontsize_col = fontsize, border_color = NA,
               treeheight_row = if (is.na(filename)) 25 else 8,
               treeheight_col = if (is.na(filename)) 25 else 8,
               legend = is.na(filename), annotation_legend = is.na(filename),
               main = main)
  if (!is.na(filename)) {
    args$filename <- file.path(FIG_DIR, filename)
    args$width    <- PUB_CM / 2.54
    args$height   <- PUB_CM / 2.54
  }
  do.call(pheatmap, args)
}

draw_heatmap(25, 2.2, "Top 25", "heatmap_top25.pdf")
```

```{r heatmap-display, fig.width=7, fig.height=9}
draw_heatmap(50, 8, "Top 50 genes by padj (SV-corrected VST, row z-score)")
```

## Osmosensitive gene panel

The gene set from Panel B of the reference figure. Bars are log2FC with
posterior SD; fill is the observed direction at `padj < 0.05`, stars are
`***` padj < 0.001, `**` < 0.01, `*` < 0.05.

```{r osmo-panel, fig.width=7, fig.height=5}
osmo_genes <- c("Hspa1b", "Sgk1", "Slc5a3", "Lrrc8a", "Slc38a2",
                "Slc12a5", "Slc6a6", "Aqp1", "Akr1b1", "Aqp4", "Nfat5")

osmo_df <- res_tbl %>%
  dplyr::filter(symbol %in% osmo_genes, !is.na(log2FC), !is.na(padj)) %>%
  group_by(symbol) %>%
  slice_min(padj, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    dir   = case_when(padj >= panel_padj_cut ~ "NS",
                      log2FC > 0 ~ "Up", TRUE ~ "Down"),
    stars = case_when(padj < 0.001 ~ "***", padj < 0.01 ~ "**",
                      padj < 0.05 ~ "*", TRUE ~ "")
  ) %>%
  arrange(desc(log2FC)) %>%
  mutate(symbol = factor(symbol, levels = symbol))

make_osmo_panel <- function(pub) {
  s <- pub_style(pub)
  ggplot(osmo_df, aes(symbol, log2FC, fill = dir)) +
    geom_col(width = s$bar_w, linewidth = s$stroke) +
    geom_errorbar(aes(ymin = log2FC - lfcSE, ymax = log2FC + lfcSE),
                  width = s$err_w, linewidth = s$stroke) +
    geom_text(aes(label = stars,
                  y = log2FC + sign(log2FC) * lfcSE,
                  vjust = ifelse(log2FC >= 0, -0.25, 1.15)),
              size = s$lab * 1.15) +
    geom_hline(yintercept = 0, color = "grey30", linewidth = s$stroke) +
    scale_fill_manual(values = sig_cols,
                      name = paste0("padj < ", panel_padj_cut)) +
    scale_y_continuous(expand = expansion(mult = c(0.16, 0.16))) +
    labs(title = "Osmosensitive genes", x = NULL,
         y = expression(log[2]~"(KO / WT)")) +
    theme_pub(pub) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"),
          panel.grid.major.x = element_blank(),
          legend.position = if (pub) "none" else "right")
}

show_both(make_osmo_panel, "panel_osmosensitive.pdf")
```

## Müller glia gliosis signature

Reactive gliosis is a **bidirectional** shift: Müller glia up-regulate a stress
and intermediate-filament program while down-regulating the homeostatic
machinery that defines the resting cell. Showing only the induced half
overstates the case, so both tiers are plotted together — the signature is
convincing when the left panel rises *and* the right panel falls.

Bars are log2FC with posterior SD; stars are `***` padj < 0.001,
`**` < 0.01, `*` < 0.05. Fill encodes the actual result (red up, blue down,
grey n.s.), which is deliberately independent of the tier a gene was assigned
to — a homeostatic marker coming out red is visible immediately.

```{r gliosis-panel, fig.width=11, fig.height=5.5}
gliosis_sets <- list(
  # Induced in reactive Müller glia: intermediate filaments, acute-phase and
  # stress response, JAK/STAT and cytokine signaling.
  `Reactive (expected up)` = c(
    "Gfap", "Vim", "Nes", "Serpina3n", "Lcn2", "Edn2", "Stat3", "Socs3",
    "A2m", "Timp1", "Cd44", "Tnc", "Chi3l1", "Osmr", "Steap4", "Ccl2",
    "C3", "Cebpd", "Hspb1", "Fgf2", "Clu", "Igfbp5", "Crim1"),
  # Homeostatic Müller identity: visual-cycle, glutamate and potassium
  # handling. Expected to fall as the cells de-differentiate.
  `Homeostatic (expected down)` = c(
    "Rlbp1", "Glul", "Slc1a3", "Aqp4", "Kcnj10", "Car14", "Dkk3",
    "Slc38a3", "Glud1", "Sox9", "Apoe")
)

gliosis_df <- tibble(tier = names(gliosis_sets), symbol = gliosis_sets) %>%
  unnest(symbol) %>%
  inner_join(res_tbl %>%
               dplyr::select(symbol, log2FC, lfcSE, padj, baseMean),
             by = "symbol") %>%
  dplyr::filter(!is.na(log2FC), !is.na(padj)) %>%
  mutate(
    tier = factor(tier, levels = names(gliosis_sets)),
    dir  = case_when(padj >= panel_padj_cut ~ "NS",
                     log2FC > 0 ~ "Up", TRUE ~ "Down"),
    stars = case_when(padj < 0.001 ~ "***", padj < 0.01 ~ "**",
                      padj < 0.05 ~ "*", TRUE ~ "")
  ) %>%
  arrange(tier, desc(log2FC)) %>%
  mutate(symbol = factor(symbol, levels = unique(symbol)))

make_gliosis_panel <- function(pub) {
  s <- pub_style(pub)
  ggplot(gliosis_df, aes(symbol, log2FC, fill = dir)) +
    geom_col(width = s$bar_w, linewidth = s$stroke) +
    geom_errorbar(aes(ymin = log2FC - lfcSE,
                      ymax = log2FC + lfcSE),
                  width = s$err_w, linewidth = s$stroke) +
    geom_text(aes(label = stars,
                  y = log2FC + sign(log2FC) * lfcSE,
                  vjust = ifelse(log2FC >= 0, -0.25, 1.15)),
              size = s$lab * 1.15) +
    geom_hline(yintercept = 0, color = "grey30", linewidth = s$stroke) +
    scale_fill_manual(values = sig_cols, name = paste0("padj < ", panel_padj_cut)) +
    scale_y_continuous(expand = expansion(mult = c(0.16, 0.16))) +
    facet_grid(~ tier, scales = "free_x", space = "free_x") +
    labs(title = "Müller glia gliosis signature",
         x = NULL, y = expression(log[2]~"(KO / WT)")) +
    theme_pub(pub) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"),
          panel.grid.major.x = element_blank(),
          legend.position = if (pub) "none" else "right")
}

# Wider than square: ~35 genes cannot share a 4 cm axis legibly.
save_pub(make_gliosis_panel(TRUE), "panel_gliosis.pdf", cm = 9, cm_h = 4.5)
make_gliosis_panel(FALSE)
```

```{r gliosis-stats}
# Is the tier as a whole shifted, beyond the individual genes? A Wilcoxon test
# against all other tested genes is a competitive test: it asks whether these
# genes are more extreme than the transcriptome background, not merely whether
# they are nonzero.
background_lfc <- res_tbl %>%
  dplyr::filter(!is.na(log2FC), !symbol %in% gliosis_df$symbol) %>%
  pull(log2FC)

gliosis_stats <- gliosis_df %>%
  group_by(tier) %>%
  summarise(
    n_genes    = n(),
    median_lfc = median(log2FC),
    n_sig_up   = sum(dir == "Up"),
    n_sig_down = sum(dir == "Down"),
    wilcox_p   = wilcox.test(log2FC, background_lfc)$p.value,
    .groups = "drop"
  )

knitr::kable(gliosis_stats, digits = c(0, 0, 3, 0, 0, 6),
             caption = "Gliosis tiers vs. transcriptome background (Wilcoxon)")
```

## Retinal cell-type marker panels

Whole-retina RNA-seq cannot distinguish a change in transcription within cells
from a change in the cellular composition of the tissue. Coordinated movement
of an entire marker panel argues for composition.

```{r retina-markers, fig.width=9, fig.height=6}
retina_markers <- list(
  Rod         = c("Rho", "Nrl", "Nr2e3", "Gnat1", "Pde6b", "Sag", "Cnga1", "Rp1"),
  Cone        = c("Opn1sw", "Opn1mw", "Arr3", "Gnat2", "Pde6h", "Gngt2", "Elovl2"),
  Bipolar     = c("Vsx2", "Grm6", "Cabp5", "Prkca", "Trpm1", "Isl1"),
  Amacrine    = c("Slc6a9", "Gad1", "Gad2", "Tfap2b", "Chat", "Slc32a1"),
  Horizontal  = c("Lhx1", "Calb1", "Onecut1", "Onecut2"),
  RGC         = c("Rbpms", "Sncg", "Nefl", "Pou4f1", "Thy1", "Slc17a6"),
  Muller      = c("Rlbp1", "Glul", "Slc1a3", "Apoe", "Sox9", "Clu", "Aqp4"),
  Astrocyte   = c("Gfap", "Pax2", "Aldh1l1"),
  Microglia   = c("Aif1", "Cx3cr1", "P2ry12", "Csf1r", "Tmem119", "C1qa"),
  RPE         = c("Rpe65", "Ttr", "Best1", "Lrat"),
  Endothelial = c("Cldn5", "Pecam1", "Flt1")
)

marker_df <- tibble(cell_type = names(retina_markers), symbol = retina_markers) %>%
  unnest(symbol) %>%
  inner_join(res_tbl %>%
               dplyr::select(symbol, log2FC, padj, baseMean), by = "symbol") %>%
  dplyr::filter(!is.na(log2FC), !is.na(padj)) %>%
  mutate(dir = case_when(padj >= panel_padj_cut ~ "NS",
                         log2FC > 0 ~ "Up", TRUE ~ "Down"),
         cell_type = factor(cell_type, levels = names(retina_markers)))

make_marker_plot <- function(pub) {
  s <- pub_style(pub)
  p <- ggplot(marker_df, aes(log2FC, fct_rev(cell_type))) +
    geom_vline(xintercept = 0, color = "grey40", linewidth = s$stroke) +
    geom_jitter(aes(color = dir), height = 0.18,
                size = s$pt * 1.5, alpha = s$alpha + 0.15) +
    scale_color_manual(values = sig_cols, name = paste0("padj < ", panel_padj_cut)) +
    labs(title = "Retinal cell-type markers",
         x = expression(log[2]~"(KO / WT)"), y = NULL) +
    theme_pub(pub) +
    theme(panel.grid.major.y = element_blank(),
          legend.position = if (pub) "none" else "right")

  if (!pub) {
    p <- p + ggrepel::geom_text_repel(
      data = marker_df %>% dplyr::filter(dir != "NS"),
      aes(label = symbol), size = s$lab, fontface = "italic",
      max.overlaps = 25, box.padding = 0.25, segment.color = "grey60")
  }
  p
}

show_both(make_marker_plot, "panel_retina_markers.pdf")
```

```{r marker-summary}
marker_summary <- marker_df %>%
  group_by(cell_type) %>%
  summarise(n_genes = n(), median_lfc = median(log2FC),
            n_up = sum(dir == "Up"), n_down = sum(dir == "Down"),
            .groups = "drop") %>%
  arrange(desc(abs(median_lfc)))

knitr::kable(marker_summary, digits = 3, caption = "Cell-type marker summary")
```

## Gene set enrichment (GSEA)

Genes ranked by log2FC. GSEA uses the whole ranked list and needs no
significance cutoff, which makes it sensitive to coordinated shifts that never
reach per-gene significance.

```{r msigdb-helper}
# msigdbr changed argument names in v10 (category/subcategory ->
# collection/subcollection). Handle both.
get_msig <- function(collection, subcollection = NULL) {
  fmls <- names(formals(msigdbr::msigdbr))
  args <- list(species = "Mus musculus")
  if ("collection" %in% fmls) {
    args$collection <- collection
    if (!is.null(subcollection)) args$subcollection <- subcollection
  } else {
    args$category <- collection
    if (!is.null(subcollection)) args$subcategory <- subcollection
  }
  do.call(msigdbr::msigdbr, args)
}

msig_to_list <- function(df) {
  df <- df %>%
    dplyr::filter(!is.na(gene_symbol), nzchar(trimws(gene_symbol))) %>%
    dplyr::distinct(gs_name, gene_symbol)
  split(df$gene_symbol, df$gs_name)
}

gene_sets <- tryCatch({
  list(Hallmark = msig_to_list(get_msig("H")),
       `GO:BP`  = msig_to_list(get_msig("C5", "GO:BP")),
       Reactome = msig_to_list(get_msig("C2", "CP:REACTOME")))
}, error = function(e) {
  message("MSigDB retrieval failed (", conditionMessage(e), "); skipping GSEA.")
  NULL
})
```

```{r gsea, fig.width=10, fig.height=12}
# fgsea rejects empty, NA and duplicated names, and a non-finite stat poisons
# the enrichment score. Enforce all four here.
ranks <- res_tbl %>%
  dplyr::filter(!is.na(symbol), nzchar(trimws(symbol)),
                !is.na(log2FC), is.finite(log2FC), !is.na(padj)) %>%
  group_by(symbol) %>%
  slice_min(padj, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  { setNames(.$log2FC, .$symbol) }

ranks <- ranks[!duplicated(names(ranks))]
ranks <- sort(ranks, decreasing = TRUE)

gsea_all <- if (!is.null(gene_sets)) {
  map_dfr(names(gene_sets), function(coll) {
    fgsea::fgsea(pathways = gene_sets[[coll]], stats = ranks,
                 minSize = 15, maxSize = 500, eps = 0) %>%
      as_tibble() %>%
      dplyr::filter(!is.na(padj)) %>%
      mutate(collection = coll)
  })
} else NULL

tidy_name <- function(x) {
  x %>% str_replace_all("^(HALLMARK|GOBP|REACTOME|WP)_", "") %>%
    str_replace_all("_", " ") %>% str_to_sentence() %>% str_trunc(48)
}

if (!is.null(gsea_all) && nrow(gsea_all) > 0) {
  write_csv(
    gsea_all %>% mutate(leadingEdge = map_chr(leadingEdge, ~ paste(.x, collapse = ";"))),
    "gsea_KOvWT_SVA.csv")

  make_gsea_plot <- function(pub, coll, n_show = 10) {
    s  <- pub_style(pub)
    df <- gsea_all %>%
      dplyr::filter(collection == coll) %>%
      slice_min(padj, n = n_show, with_ties = FALSE) %>%
      mutate(pathway = tidy_name(pathway), dir = ifelse(NES > 0, "Up", "Down"))
    if (nrow(df) == 0) return(NULL)

    ggplot(df, aes(NES, reorder(pathway, NES), fill = dir)) +
      geom_col(width = s$bar_w, linewidth = s$stroke) +
      geom_vline(xintercept = 0, color = "grey30", linewidth = s$stroke) +
      scale_fill_manual(values = sig_cols, name = NULL) +
      labs(title = paste0("GSEA: ", coll), x = "NES", y = NULL) +
      theme_pub(pub) +
      theme(panel.grid.major.y = element_blank(),
            axis.text.y = element_text(size = rel(if (pub) 0.7 else 0.9)),
            legend.position = if (pub) "none" else "right")
  }

  for (coll in names(gene_sets)) {
    p <- make_gsea_plot(TRUE, coll, n_show = 8)
    if (!is.null(p)) {
      save_pub(p, paste0("gsea_", str_replace(coll, ":", ""), ".pdf"),
               cm = 8, cm_h = 4.5)
    }
  }

  gsea_display <- compact(map(names(gene_sets), ~ make_gsea_plot(FALSE, .x, 12)))
  if (length(gsea_display) > 0) wrap_plots(gsea_display, ncol = 1)
}
```

## Pathway over-representation analysis

GSEA and over-representation answer different questions. GSEA asks whether a
gene set drifts systematically across the whole ranked list; ORA asks whether
the `padj < 0.01` set is enriched for a pathway relative to the tested
background. A result supported by both is considerably more secure than one
found by either alone.

Four databases are queried, up- and down-regulated genes separately. KEGG and
WikiPathways fetch over the network at render time, so each call is wrapped —
a failure logs a message and drops that database rather than killing the
render.

```{r ora-setup}
# The universe is every gene that survived filtering and has an Entrez ID.
# Using all of Entrez instead would inflate every enrichment, since genes not
# expressed in retina could never have been called.
universe_entrez <- res_tbl %>%
  dplyr::filter(!is.na(entrez), nzchar(trimws(entrez)), !is.na(padj)) %>%
  pull(entrez) %>% unique()

sig_entrez <- function(direction) {
  sig_tbl %>%
    dplyr::filter(!is.na(entrez), nzchar(trimws(entrez)),
                  if (direction == "up") log2FC > 0 else log2FC < 0) %>%
    pull(entrez) %>% unique()
}

cat("Universe:", length(universe_entrez), "genes\n")
cat("Up:", length(sig_entrez("up")), " Down:", length(sig_entrez("down")), "\n")

# Each entry returns an enrichResult or NULL. GO is simplified to collapse
# redundant parent/child terms, which otherwise dominate the top of the table.
ora_specs <- list(
  `GO:BP` = function(g, u) {
    e <- clusterProfiler::enrichGO(g, OrgDb = org.Mm.eg.db, universe = u,
                                   keyType = "ENTREZID", ont = "BP",
                                   pAdjustMethod = "BH", qvalueCutoff = 0.05,
                                   readable = TRUE)
    if (!is.null(e) && nrow(as.data.frame(e)) > 0) {
      e <- clusterProfiler::simplify(e, cutoff = 0.7, by = "p.adjust", select_fun = min)
    }
    e
  },
  `GO:MF` = function(g, u) {
    e <- clusterProfiler::enrichGO(g, OrgDb = org.Mm.eg.db, universe = u,
                                   keyType = "ENTREZID", ont = "MF",
                                   pAdjustMethod = "BH", qvalueCutoff = 0.05,
                                   readable = TRUE)
    if (!is.null(e) && nrow(as.data.frame(e)) > 0) {
      e <- clusterProfiler::simplify(e, cutoff = 0.7, by = "p.adjust", select_fun = min)
    }
    e
  },
  `GO:CC` = function(g, u) {
    e <- clusterProfiler::enrichGO(g, OrgDb = org.Mm.eg.db, universe = u,
                                   keyType = "ENTREZID", ont = "CC",
                                   pAdjustMethod = "BH", qvalueCutoff = 0.05,
                                   readable = TRUE)
    if (!is.null(e) && nrow(as.data.frame(e)) > 0) {
      e <- clusterProfiler::simplify(e, cutoff = 0.7, by = "p.adjust", select_fun = min)
    }
    e
  },
  KEGG = function(g, u) {
    e <- clusterProfiler::enrichKEGG(g, organism = "mmu", keyType = "kegg",
                                     universe = u, pAdjustMethod = "BH",
                                     qvalueCutoff = 0.05)
    if (!is.null(e) && nrow(as.data.frame(e)) > 0) {
      e <- DOSE::setReadable(e, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
    }
    e
  },
  Reactome = function(g, u) {
    ReactomePA::enrichPathway(g, organism = "mouse", universe = u,
                              pAdjustMethod = "BH", qvalueCutoff = 0.05,
                              readable = TRUE)
  },
  WikiPathways = function(g, u) {
    e <- clusterProfiler::enrichWP(g, organism = "Mus musculus", universe = u,
                                   pAdjustMethod = "BH", qvalueCutoff = 0.05)
    if (!is.null(e) && nrow(as.data.frame(e)) > 0) {
      e <- DOSE::setReadable(e, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
    }
    e
  }
)

# GeneRatio and BgRatio arrive as "12/300" strings; fold enrichment is the
# ratio of the two, and is what makes results comparable across databases with
# very different pathway sizes.
parse_ratio <- function(x) {
  p <- str_split_fixed(as.character(x), "/", 2)
  as.numeric(p[, 1]) / as.numeric(p[, 2])
}

run_one_ora <- function(db, direction) {
  genes <- sig_entrez(direction)
  if (length(genes) < 10) {
    message("Only ", length(genes), " ", direction,
            " genes with Entrez IDs; skipping ", db, ".")
    return(tibble())
  }

  e <- tryCatch(ora_specs[[db]](genes, universe_entrez),
                error = function(err) {
                  message(db, " (", direction, ") failed: ", conditionMessage(err))
                  NULL
                })

  if (is.null(e)) return(tibble())
  df <- as.data.frame(e)
  if (nrow(df) == 0) return(tibble())

  as_tibble(df) %>%
    mutate(database = db, direction = direction,
           fold_enrichment = parse_ratio(GeneRatio) / parse_ratio(BgRatio)) %>%
    # any_of("qvalue"): enrichKEGG and enrichWP do not always return a qvalue
    # column, and a bare select() would error on the first database that omits
    # it, taking the whole suite down with it.
    dplyr::select(database, direction, ID, Description, GeneRatio, BgRatio,
                  fold_enrichment, pvalue, p.adjust, any_of("qvalue"),
                  Count, geneID)
}
```

```{r ora-run}
ora_all <- map_dfr(names(ora_specs), function(db) {
  map_dfr(c("up", "down"), function(d) run_one_ora(db, d))
})

if (nrow(ora_all) > 0) {
  write_csv(ora_all, "ora_KOvWT_SVA_allDatabases.csv")

  # Excel forbids : \ / ? * [ ] in sheet names and caps them at 31 characters,
  # so "GO:BP" would fail. Sanitize before writing.
  ora_sheets <- split(ora_all, ora_all$database)
  names(ora_sheets) <- names(ora_sheets) %>%
    str_replace_all("[:\\\\/?*\\[\\]]", "_") %>%
    str_trunc(31, ellipsis = "")

  openxlsx::write.xlsx(ora_sheets,
                       file = "ora_KOvWT_SVA_allDatabases.xlsx", overwrite = TRUE)

  ora_summary <- ora_all %>%
    dplyr::filter(p.adjust < 0.05) %>%
    count(database, direction) %>%
    pivot_wider(names_from = direction, values_from = n, values_fill = 0)

  knitr::kable(ora_summary,
               caption = "Significantly enriched terms per database (padj < 0.05)")
} else {
  message("No over-representation results returned by any database.")
}
```

```{r ora-dotplots, fig.width=11, fig.height=7}
# Standard dotplot: fold enrichment on x, gene count as size, significance as
# color. Preferred over a bare -log10(p) bar chart because a highly significant
# term drawn from a 900-gene pathway is a weaker result than a modest one from
# a 30-gene pathway, and only fold enrichment shows that.
make_ora_dot <- function(pub, db, n_show = 10) {
  s  <- pub_style(pub)
  df <- ora_all %>%
    dplyr::filter(database == db, p.adjust < 0.05) %>%
    group_by(direction) %>%
    slice_min(p.adjust, n = n_show, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(Description = str_trunc(Description, 46),
           direction = recode(direction, up = "Up in KO", down = "Down in KO"),
           direction = factor(direction, levels = c("Up in KO", "Down in KO")))
  if (nrow(df) == 0) return(NULL)

  ggplot(df, aes(fold_enrichment, reorder(Description, fold_enrichment))) +
    geom_point(aes(size = Count, color = -log10(p.adjust))) +
    scale_color_viridis_c(option = "plasma", end = 0.9,
                          name = expression(-log[10]~padj)) +
    scale_size_continuous(range = if (pub) c(0.4, 1.8) else c(1.5, 6),
                          name = "Genes") +
    facet_wrap(~ direction, scales = "free_y", ncol = 1) +
    labs(title = db, x = "Fold enrichment", y = NULL) +
    theme_pub(pub) +
    theme(panel.grid.major.y = element_line(linewidth = s$stroke * 0.5),
          axis.text.y = element_text(size = rel(if (pub) 0.65 else 0.9)),
          legend.position = if (pub) "none" else "right")
}

if (nrow(ora_all) > 0) {
  for (db in unique(ora_all$database)) {
    p <- make_ora_dot(TRUE, db, n_show = 6)
    if (!is.null(p)) {
      save_pub(p, paste0("ora_", str_replace_all(db, "[: ]", ""), ".pdf"),
               cm = 8, cm_h = 6)
    }
  }

  ora_display <- compact(map(unique(ora_all$database), ~ make_ora_dot(FALSE, .x, 10)))
  if (length(ora_display) > 0) wrap_plots(ora_display, ncol = 2)
}
```

```{r ora-top-table}
if (nrow(ora_all) > 0) {
  knitr::kable(
    ora_all %>%
      dplyr::filter(p.adjust < 0.05) %>%
      group_by(database, direction) %>%
      slice_min(p.adjust, n = 3, with_ties = FALSE) %>%
      ungroup() %>%
      dplyr::select(database, direction, Description, fold_enrichment,
                    Count, p.adjust),
    digits = c(0, 0, 0, 2, 0, 6),
    caption = "Top 3 enriched terms per database and direction")
}
```

### Where GSEA and ORA agree

```{r gsea-ora-overlap}
# Matching on cleaned term names is approximate -- the same biology is named
# differently across databases -- so treat this as a pointer, not a tally.
if (!is.null(gsea_all) && nrow(gsea_all) > 0 && nrow(ora_all) > 0) {
  norm_term <- function(x) {
    x %>% str_replace_all("^(HALLMARK|GOBP|REACTOME|WP)_", "") %>%
      str_replace_all("[_[:punct:]]", " ") %>% str_squish() %>% str_to_lower()
  }

  gsea_sig <- gsea_all %>%
    dplyr::filter(padj < 0.05) %>%
    mutate(term = norm_term(pathway),
           gsea_dir = ifelse(NES > 0, "up", "down")) %>%
    dplyr::select(term, gsea_dir, NES, gsea_padj = padj)

  ora_sig <- ora_all %>%
    dplyr::filter(p.adjust < 0.05) %>%
    mutate(term = norm_term(Description)) %>%
    dplyr::select(term, ora_dir = direction, database,
                  fold_enrichment, ora_padj = p.adjust)

  overlap <- inner_join(gsea_sig, ora_sig, by = "term", relationship = "many-to-many") %>%
    dplyr::filter(gsea_dir == ora_dir) %>%
    arrange(gsea_padj) %>%
    distinct(term, .keep_all = TRUE)

  if (nrow(overlap) > 0) {
    knitr::kable(head(overlap, 20), digits = 3,
                 caption = "Terms called by both GSEA and ORA, same direction")
  } else {
    message("No exact term-name matches between GSEA and ORA outputs.")
  }
}
```

## Counts for individual genes

```{r plot-counts, fig.width=10, fig.height=7}
# A gene driven by a single sample is obvious here and nowhere else.
make_counts_plot <- function(pub, n_genes = 9) {
  s   <- pub_style(pub)
  top <- sig_tbl %>% slice_min(padj, n = n_genes, with_ties = FALSE)
  if (nrow(top) == 0) return(NULL)

  counts_df <- map_dfr(seq_len(nrow(top)), function(i) {
    d <- plotCounts(dds, gene = top$ensemblID[i], intgroup = "background",
                    normalized = TRUE, returnData = TRUE)
    d$gene <- ifelse(is.na(top$symbol[i]), top$ensemblID[i], top$symbol[i])
    d
  })

  ggplot(counts_df, aes(background, count, color = background)) +
    geom_point(position = position_jitter(width = 0.12),
               size = s$pt * 2, alpha = 0.9) +
    scale_color_manual(values = geno_cols, guide = "none") +
    scale_y_log10(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
    facet_wrap(~ gene, scales = "free_y", ncol = 3) +
    labs(title = "Top genes", x = NULL, y = "Norm. count") +
    theme_pub(pub) +
    theme(strip.text = element_text(face = "italic", size = rel(0.8)))
}

p <- make_counts_plot(TRUE, 6)
if (!is.null(p)) save_pub(p, "counts_top_genes.pdf", cm = 6, cm_h = 6)
make_counts_plot(FALSE, 9)
```

## Appendix: why SVA

Revision 4 ran every analysis twice, with and without the surrogate variable,
to test whether the adjustment was absorbing genotype signal rather than noise.
The comparison is condensed here as justification for the modeling choice.

```{r appendix-nosva}
dds_nosva <- dds_pre
design(dds_nosva) <- ~ background
dds_nosva <- DESeq(dds_nosva, quiet = TRUE)

res_nosva_mle <- results(dds_nosva, name = "background_KO_vs_WT", alpha = padj_cut)
res_nosva <- lfcShrink(dds_nosva, coef = "background_KO_vs_WT",
                       type = "apeglm", res = res_nosva_mle, quiet = TRUE)

compare_tbl <- tibble(
  ensemblID   = rownames(res_nosva),
  noSVA_log2FC = res_nosva$log2FoldChange,
  noSVA_padj   = res_nosva$padj
) %>%
  full_join(res_tbl %>%
              dplyr::select(ensemblID, symbol, baseMean,
                            SVA_log2FC = log2FC, SVA_padj = padj),
            by = "ensemblID") %>%
  mutate(called = case_when(
    noSVA_padj < padj_cut & SVA_padj < padj_cut    ~ "Both",
    noSVA_padj < padj_cut & !(SVA_padj < padj_cut) ~ "No SVA only",
    SVA_padj < padj_cut & !(noSVA_padj < padj_cut) ~ "SVA only",
    TRUE ~ "Neither"))

write_csv(compare_tbl, "appendix_SVA_comparison.csv")

knitr::kable(
  tibble(
    model = c("No SVA", "SVA"),
    n_sig = c(sum(compare_tbl$noSVA_padj < padj_cut, na.rm = TRUE),
              sum(compare_tbl$SVA_padj < padj_cut, na.rm = TRUE))
  ),
  caption = paste0("Genes called at padj < ", padj_cut))

knitr::kable(count(compare_tbl, called),
             caption = "Agreement between the two models")
```

```{r appendix-scatter, fig.width=6, fig.height=5}
# Systematic compression toward zero on the SVA axis would be the signature of
# a surrogate variable that had absorbed part of the genotype effect. A slope
# at or near 1 says it did not.
p_appendix_fun <- function(pub) {
  s  <- pub_style(pub)
  df <- compare_tbl %>% dplyr::filter(!is.na(noSVA_log2FC), !is.na(SVA_log2FC))
  r     <- round(cor(df$noSVA_log2FC, df$SVA_log2FC), 3)
  slope <- round(coef(lm(SVA_log2FC ~ noSVA_log2FC, data = df))[2], 3)

  ggplot(df, aes(noSVA_log2FC, SVA_log2FC)) +
    geom_point(alpha = s$alpha * 0.6, size = s$pt, color = "#3B4A6B") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "grey40", linewidth = s$stroke) +
    geom_smooth(method = "lm", se = FALSE, color = col_up,
                linewidth = s$stroke * 1.6, formula = y ~ x) +
    labs(title = "log2FC: SVA vs. no SVA",
         subtitle = paste0("r = ", r, ", slope = ", slope),
         x = "No SVA", y = "SVA") +
    theme_pub(pub)
}

show_both(p_appendix_fun, "appendix_lfc_scatter.pdf")
```

## Figure inventory

```{r figure-inventory}
fig_files <- list.files(FIG_DIR, pattern = "\\.pdf$", full.names = TRUE)

cat("Figures written to '", FIG_DIR, "/'. Square panels are ",
    PUB_CM, " x ", PUB_CM, " cm; multi-panel and pathway figures are wider ",
    "where a 4 cm axis cannot hold the labels legibly.\n", sep = "")

knitr::kable(
  tibble(file = basename(fig_files),
         size_kb = round(file.size(fig_files) / 1024, 1)) %>% arrange(file),
  caption = "Publication figures")
```

## Session info

```{r session-info}
sessionInfo()
```
