---
title: "Supplemental File S1:"
subtitle: "Statistical outputs for OCT data (Tables S1A-S4C)"
format:
  pdf:
    pdf-engine: lualatex
    toc: true
    toc-depth: 3
    number-sections: false
    colorlinks: true
    linkcolor: navy
    urlcolor: navy
    fig-pos: "H"
    fig-align: center
    keep-tex: false
    highlight-style: github
    fontsize: 11pt
    mainfont: "TeX Gyre Heros"
    sansfont: "TeX Gyre Heros"
    monofont: "DejaVu Sans Mono"
    geometry:
      - top=1in
      - bottom=1in
      - left=1in
      - right=1in
    include-in-header:
      text: |
        \usepackage[table]{xcolor}
        \usepackage{colortbl}
        \usepackage{setspace}
        \setstretch{1.15}
        \usepackage{float}
        \usepackage{booktabs}
        \usepackage{threeparttable}
        \usepackage{tabularx}
        \usepackage{makecell}
        \usepackage{caption}
        \captionsetup[table]{labelformat=empty,labelsep=none}
        \usepackage{float}
        \floatplacement{table}{H}

          
execute:
  echo: false        
  warning: false
  message: false
  error: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(purrr)
library(stringr)

library(readxl)
library(knitr)
library(kableExtra)
library(broom.mixed)

library(lme4)     
library(lmerTest) 
library(emmeans)  
library(ggpubr)   
library(ggsignif)
library(patchwork)

```

```{r}
#| label: table-helper1
#| include: true
#| echo: false

iovs_makecell_headers <- function(nms) {
  # 1) replace <br> or <BR> with LaTeX line breaks
  nms <- gsub("(?i)<br\\s*/?>", " \\\\ ", nms, perl = TRUE)
  # 2) wrap any header containing '\\' in \makecell{...}
  nms <- vapply(nms, function(x) {
    if (grepl("\\\\\\\\", x)) sprintf("\\makecell{%s}", x) else x
  }, character(1))
  nms
}

tab_iovs <- function(df, caption = NULL, align = NULL, font_size = 10) {
  if (is.null(align)) {
    align <- paste(ifelse(vapply(df, is.numeric, TRUE), "r", "l"), collapse = "")
  }

  
  colnames(df) <- iovs_makecell_headers(colnames(df))

  knitr::kable(
    df,
    format    = "latex",
    booktabs  = TRUE,  
    caption   = caption,
    align     = align,
    escape    = FALSE,  
    longtable = FALSE,  
    linesep   = ""      
  ) |>
    kableExtra::kable_styling(
      latex_options = c("hold_position"),
      font_size     = font_size
    )
}

```

```{r}
#| label: table-helper2
#| include: true
#| echo: false

tidy_anova <- function(anova_obj, digits = 2, p_digits = 2) {
  df <- tryCatch(broom::tidy(anova_obj), error = function(e) NULL)
  if (is.null(df)) {
    df <- as.data.frame(anova_obj)
    if (!("term" %in% names(df))) {
      df <- tibble::rownames_to_column(df, var = "term")
    }
  }

  nm <- names(df)
  has_numden <- all(c("NumDF","DenDF") %in% nm)

  out <- if (has_numden) {
    df |>
      dplyr::select(dplyr::any_of(c("term","sumsq","meansq","NumDF","DenDF","statistic","p.value"))) |>
      dplyr::rename(
        Term       = term,
        `Sum Sq`   = sumsq,
        `Mean Sq`  = meansq,
        `F value`  = statistic,
        `Pr(>F)`   = p.value
      )
  } else {
    df |>
      dplyr::select(dplyr::any_of(c("term","sumsq","meansq","df","statistic","p.value"))) |>
      dplyr::rename(
        Term       = term,
        `Sum Sq`   = sumsq,
        `Mean Sq`  = meansq,
        DF         = df,
        `F value`  = statistic,
        `Pr(>F)`   = p.value
      )
  }

  out |>
    dplyr::mutate(
      dplyr::across(where(is.numeric) & !dplyr::matches("^Pr\\(>F\\)$"),
                    ~ round(.x, digits)),
      `Pr(>F)` = if ("Pr(>F)" %in% names(out))
                   formatC(`Pr(>F)`, format = "e", digits = p_digits)
                 else NULL
    )
}

```

## Method

We used a linear mixed model with fixed effects of genotype, condition, side, and all interactions. The model included a random intercept for mouse nested within genotype. The test of interest was the interaction between genotype and condition. We used the model to estimate means for all conditions.

## 1. ELM-RPE

```{r}
#| label: elm-rpe_modeling
#| include: true
#| echo: false
#| warning: false


data_er <- read_excel("Kv82_exp7_all_biomarkers.xlsx", sheet = "elmrpe")
data_er_long <- data_er |>
    pivot_longer(
    cols = c(light, dark),
    names_to = "condition",
    values_to = "value")

# Per-mouse averages preserving genotype × CONDITION
data_er_c <- data_er_long |>
  group_by(mouse, genotype, condition) |>
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") |>
  mutate(
    genotype    = factor(genotype, levels = c("Het", "KO")),
    condition = factor(condition, levels = c("dark", "light"),
                       labels = c("Dark", "Light"))
  )

# Per-mouse averages preserving genotype × SIDE
data_er_s <- data_er_long |>
  group_by(mouse, genotype, side) |>
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") |>
  mutate(
    genotype = factor(genotype, levels = c("Het", "KO")),
    side   = factor(side,   levels = c("Inferior", "Superior"))
  )
# Model 1: genotype × Condition
model_er_cond <- lmer(mean_value ~ genotype * condition + (1 | mouse), data = data_er_c)
anova_er_cond <- anova(model_er_cond, type = 3)
anova_er_cond_tidy <- tidy_anova(anova_er_cond)
tab_iovs(anova_er_cond_tidy, 
         caption = "\\textbf{Table S1A. }ELM-RPE: Type III ANOVA results (genotype × condition)")


# Model 2: genotype × Side
model_er_side <- lmer(mean_value ~ genotype * side + (1 | mouse), data = data_er_s)
anova_er_side <- anova(model_er_side, type = 3)
anova_er_side_tidy <- tidy_anova(anova_er_side)
tab_iovs(anova_er_side_tidy, 
         caption = "\\textbf{Table S1B. }ELM-RPE: Type III ANOVA results (genotype × side)")


```

-   The main effect of genotype on ELM-RPE thickness was highly significant (p \< 0.0001).

-   There were also significant effects of condition (p \< 0.0001), the genotype x condition interaction (p = 0.0010), side (p \< 0.0001) and the genotype x side interaction (p = 0.007). Therefore, data were analyzed separately by condition and side.

```{r}
#| label: tbl_elm-rpe_emms
#| include: true
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "\\textbf{Table S1C. } ELM–RPE: Estimated marginal means."


data_er_long <- data_er |>
  tidyr::pivot_longer(c(light, dark), names_to = "condition", values_to = "value") |>
  dplyr::mutate(
    genotype    = factor(genotype,   levels = c("Het","KO")),
    condition = factor(condition, levels = c("dark","light"), labels = c("Dark","Light")),
    side      = factor(side,     levels = c("Inferior","Superior")),
    mouse     = factor(mouse)
  )


er_lmm_model <- lmer(value ~ genotype * condition * side + (1 | mouse), data = data_er_long)

er_emm <- emmeans(er_lmm_model, ~ genotype | condition * side)

er_emm_df <- as.data.frame(confint(er_emm, level = 0.95)) |>
  dplyr::transmute(
    Condition = condition,
    Side      = side,
    Genotype    = genotype,
    Mean      = round(emmean, 2),
    SE        = round(SE, 2),
    'lower CI'       = round(lower.CL, 2),
    'upper CI'       = round(upper.CL, 2)
  )
 
knitr::kable(
  er_emm_df,
  format    = "latex",
  booktabs  = TRUE,
  caption   = NULL,                    
  align     = c("l","l","r","r","r","r"),
  escape    = TRUE,
  longtable = FALSE,
  linesep   = ""
)
```

```{r}
#| label: tbl_elm-rpe_contrasts_genotype
#| include: true
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "\\textbf{Table S1D. }ELM–RPE: Het–KO contrasts by condition within side."

# Het–KO contrasts (Holm, with 95% CI)
er_contr_hetko <- contrast(er_emm, method = "pairwise",
                           by = c("condition","side"), adjust = "holm")
hk_sum <- as.data.frame(summary(er_contr_hetko, infer = c(TRUE, TRUE))) |>
  dplyr::transmute(
    Condition = condition,
    Side      = side,
    'Estimate' = round(estimate, 2),
    'lower CI' = round(lower.CL, 2),
    'upper CI' = round(upper.CL, 2),
    'Holm-adj. p'   = formatC(p.value, format = "e", digits = 2)
  ) |>
  dplyr::distinct(Condition, Side, .keep_all = TRUE)

knitr::kable(
  hk_sum,
  format    = "latex",
  booktabs  = TRUE,
  caption   = NULL,                    
  align     = c("l","l","r","r","r","r"),
  escape    = TRUE,
  longtable = FALSE,
  linesep   = ""
)

```

```{r}
#| label: tbl_elm-rpe_contrasts_cond
#| include: true
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "\\textbf{Table S1E. }ELM-RPE: Light–Dark contrasts by genotype within side."

# Light–Dark contrasts (Holm, with 95% CI)
er_contr_ld <- contrast(er_emm, method = "revpairwise",
                        by = c("genotype","side"), adjust = "holm")
ld_sum <- as.data.frame(summary(er_contr_ld, infer = c(TRUE, TRUE))) |>
  dplyr::transmute(
    Genotype = genotype,
    Side   = side,
    'Estimate' = round(estimate, 2),
    'lower CI' = round(lower.CL, 2),
    'upper CI' = round(upper.CL, 2),
    'Holm-adj. p'   = formatC(p.value, format = "e", digits = 2)
  )

knitr::kable(
  ld_sum,
  format    = "latex",
  booktabs  = TRUE,
  caption   = NULL,                    
  align     = c("l","l","r","r","r","r"),
  escape    = TRUE,
  longtable = FALSE,
  linesep   = ""
)

```

## 2. $\Delta$ ELM-RPE

```{r}
#| label: delta_interactions
#| include: true
#| echo: false
#| warning: false

data_er_delta <- data_er |>
  dplyr::filter(!is.na(delta)) |>
  dplyr::mutate(
    genotype = factor(genotype, levels = c("Het","KO")),
    side   = factor(side,   levels = c("Inferior","Superior"))
  ) 

er_delta_lmm <- lmerTest::lmer(delta ~ genotype * side + (1 | mouse),
                               data = data_er_delta)

# Model: genotype × Side
model_er_delta <- lmer(delta ~ genotype * side + (1 | mouse), data = data_er_delta)
anova_er_delta <- anova(model_er_delta, type = 3)
anova_er_delta_tidy <- tidy_anova(anova_er_delta)
tab_iovs(anova_er_side_tidy, caption = "\\textbf{Table S2A. }Delta ELM-RPE: Type III ANOVA Results (genotype × side).")

```

-   The main effect of genotype on ΔELM-RPE thickness was highly significant (p \< 0.0001).

-   There were also significant effects of side (p \< 0.0001) and the genotype x side interaction (p = 0.0074). Therefore, data were analyzed separately by side.

```{r}
#| label: tbl_delta_emms
#| include: true
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "\\textbf{Table S2B. }Delta ELM-RPE: Estimated marginal means."

data_delta <- data_er |>
  dplyr::mutate(
    genotype = factor(genotype, levels = c("Het","KO")),
    side   = factor(side,   levels = c("Inferior","Superior")),
    delta  = light - dark
  ) |>
  dplyr::select(mouse, genotype, side, delta)


delta_lmm <- lme4::lmer(delta ~ genotype * side + (1 | mouse), data = data_delta)
delta_emm <- emmeans(delta_lmm, ~ genotype | side)

d_emm_df <- as.data.frame(confint(delta_emm, level = 0.95)) |>
  dplyr::transmute(
    Side   = side,
    Genotype = genotype,
    Mean   = round(emmean, 2),
    SE     = round(SE, 2),
    'lower CI'    = round(lower.CL, 2),
    'upper CI'    = round(upper.CL, 2)
  ) |>
  dplyr::arrange(Side, Genotype)

knitr::kable(
  d_emm_df,
  format    = "latex",
  booktabs  = TRUE,
  caption   = NULL,                    
  align     = c("l","l","r","r","r","r"),
  escape    = TRUE,
  longtable = FALSE,
  linesep   = ""
)

```

```{r}
#| label: tbl_delta_emms
#| include: true
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "\\textbf{Table S2Bi. }Delta ELM-RPE: Estimated marginal means, with SD."

data_delta <- data_er |>
  dplyr::mutate(
    genotype = factor(genotype, levels = c("Het", "KO")),
    side     = factor(side, levels = c("Inferior", "Superior")),
    delta    = light - dark
  ) |>
  dplyr::select(mouse, genotype, side, delta)

delta_lmm <- lme4::lmer(delta ~ genotype * side + (1 | mouse), data = data_delta)

delta_emm <- emmeans::emmeans(delta_lmm, ~ genotype | side)

sd_df <- data_delta |>
  dplyr::group_by(side, genotype) |>
  dplyr::summarise(
    SD = stats::sd(delta, na.rm = TRUE),
    n  = sum(!is.na(delta)),
    .groups = "drop"
  )

d_emm_df <- as.data.frame(confint(delta_emm, level = 0.95)) |>
  dplyr::left_join(
    sd_df,
    by = c("side", "genotype")
  ) |>
  dplyr::transmute(
    Side       = side,
    Genotype   = genotype,
    Mean       = round(emmean, 2),
    SD         = round(SD, 2),
    SE         = round(SE, 2),
    `lower CI` = round(lower.CL, 2),
    `upper CI` = round(upper.CL, 2),
    n          = n
  ) |>
  dplyr::arrange(Side, Genotype)

knitr::kable(
  d_emm_df,
  format    = "latex",
  booktabs  = TRUE,
  caption   = NULL,
  align     = c("l", "l", "r", "r", "r", "r", "r", "r"),
  escape    = TRUE,
  longtable = FALSE,
  linesep   = ""
)
```

```{r}
#| label: tbl_delta_contrasts
#| include: true
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "\\textbf{Table S2C. }Delta ELM-RPE: Het–KO contrasts within side."

delta_contr_hk <- emmeans::contrast(delta_emm, method = "pairwise", by = "side", adjust = "holm")

hk_tbl <- as.data.frame(summary(delta_contr_hk, infer = c(TRUE, TRUE))) |>
  dplyr::transmute(
    Side     = side,
    Estimate = round(estimate, 2),
    'lower CI'      = round(lower.CL, 2),
    'upper CI'      = round(upper.CL, 2),
    `Holm-adj. p` = formatC(p.value, format = "e", digits = 2)
  ) |>
  dplyr::arrange(Side)

knitr::kable(
  hk_tbl,
  format    = "latex",
  booktabs  = TRUE,
  caption   = NULL,                    
  align     = c("l","r","r","r","r"),
  escape    = TRUE,
  longtable = FALSE,
  linesep   = ""
)
```

```{r}

```

## 3. ONL

```{r}
#| label: onl_tidy_4_modeling
#| include: true

data_onl <- read_excel("Kv82_exp7_all_biomarkers.xlsx", sheet = "onl")
data_onl_long <- data_onl |>
    pivot_longer(
    cols = c(light, dark),
    names_to = "condition",
    values_to = "value")

# 1) Per-mouse averages preserving genotype × CONDITION
data_onl_cond <- data_onl_long |>
  group_by(mouse, genotype, condition) |>
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") |>
  mutate(
    genotype    = factor(genotype, levels = c("Het", "KO")),
    condition = factor(condition, levels = c("dark", "light"),
                       labels = c("Dark", "Light"))
  )

# 2) Per-mouse averages preserving genotype × SIDE
data_onl_side <- data_onl_long |>
  group_by(mouse, genotype, side) |>
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") |>
  mutate(
    genotype = factor(genotype, levels = c("Het", "KO")),
    side   = factor(side,   levels = c("Superior", "Inferior"))
  )

```

```{r}
#| label: onl_interactions
#| include: true
#| echo: false
#| warning: false


# Model 1: genotype × Condition
model_onl_cond <- lmer(mean_value ~ genotype * condition + (1 | mouse), data = data_onl_cond)
anova_onl_cond <- anova(model_onl_cond, type = 3)
anova_onl_cond_tidy <- tidy_anova(anova_onl_cond)
tab_iovs(anova_onl_cond_tidy, 
      caption = "\\textbf{Table S3A. }ONL: Type III ANOVA results (genotype × condition)")


# Model 2: genotype × Side
model_onl_side <- lmer(mean_value ~ genotype * side + (1 | mouse), data = data_onl_side)
anova_onl_side <- anova(model_onl_side, type = 3)
anova_onl_side_tidy <- tidy_anova(anova_onl_side)
tab_iovs(anova_onl_side_tidy, 
        caption = "\\textbf{Table S3B. }ONL: Type III ANOVA results (genotype × side)")

```

-   The main effect of genotype on ONL thickness was highly significant (p \< 0.0001).

-   The effects of condition (p = 0.58) and side (p = 0.63) were not significant.

-   However, the genotype × condition (p=0.013) and genotype × side (p = 0.0003) interactions were significant. Therefore, data were analyzed separately by condition and side.

```{r}
#| label: tbl_onl_emms
#| include: true
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "\\textbf{Table S3C. }ONL: Estimated marginal means."

data_onl_long <- data_onl |>
  tidyr::pivot_longer(c(light, dark), names_to = "condition", values_to = "value") |>
  dplyr::mutate(
    genotype    = factor(genotype,   levels = c("Het","KO")),
    condition = factor(condition, levels = c("dark","light"),
                        labels = c("Dark","Light")),
    side      = factor(side,     levels = c("Inferior","Superior")),
    mouse     = factor(mouse)
  )

# Type-III contrasts
options(contrasts = c("contr.sum", "contr.poly"))

onl_lmm_model <- lmer(value ~ genotype * condition * side + (1 | mouse),
                      data = data_onl_long)

onl_emm <- emmeans(onl_lmm_model, ~ genotype | condition * side)

# Format like Table S3
onl_emm_df <- as.data.frame(summary(onl_emm, infer = TRUE)) |>
  dplyr::select(condition, side, genotype, emmean, SE, lower.CL, upper.CL) |>
  dplyr::rename(
    Condition = condition,
    Side      = side,
    Genotype    = genotype,
    Mean      = emmean,
    `lower CI`= lower.CL,
    `upper CI`= upper.CL
  ) |>
  dplyr::mutate(
    Mean     = round(Mean, 2),
    SE       = round(SE, 2),
    `lower CI` = round(`lower CI`, 2),
    `upper CI` = round(`upper CI`, 2)
  )

knitr::kable(
  onl_emm_df,
  format    = "latex",
  booktabs  = TRUE,
  align     = c("l","l","l","r","r","r","r"),
  escape    = TRUE,
  longtable = FALSE,
  linesep   = ""
)

```

```{r}
#| label: tbl_onl_contrasts_genotype
#| include: true
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "\\textbf{Table S3D. }ONL: Het–KO contrasts by condition within side."

# Het–KO contrasts (Holm, with 95% CI)
onl_contr_hetko <- contrast(onl_emm, method = "pairwise",
                           by = c("condition","side"), adjust = "holm")
onl_hk_sum <- as.data.frame(summary(onl_contr_hetko, infer = c(TRUE, TRUE))) |>
  dplyr::transmute(
    Condition = condition,
    Side      = side,
    'Estimate' = round(estimate, 2),
    'lower CI' = round(lower.CL, 2),
    'upper CI' = round(upper.CL, 2),
    'Holm-adj. p'   = formatC(p.value, format = "e", digits = 2)
  ) |>
  dplyr::distinct(Condition, Side, .keep_all = TRUE)

knitr::kable(
  onl_hk_sum,
  format    = "latex",
  booktabs  = TRUE,
  caption   = NULL,                    
  align     = c("l","l","r","r","r","r"),
  escape    = TRUE,
  longtable = FALSE,
  linesep   = ""
)

```

```{r}
#| label: tbl_onl_contrasts_cond
#| include: true
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "\\textbf{Table S3E. }ONL: Light–Dark contrasts by genotype within side."

# Light–Dark contrasts (Holm, with 95% CI)
onl_contr_ld <- contrast(onl_emm, method = "revpairwise",
                        by = c("genotype","side"), adjust = "holm")
onl_ld_sum <- as.data.frame(summary(onl_contr_ld, infer = c(TRUE, TRUE))) |>
  dplyr::transmute(
    Genotype = genotype,
    Side   = side,
    'Estimate' = round(estimate, 2),
    'lower CI' = round(lower.CL, 2),
    'upper CI' = round(upper.CL, 2),
    'Holm-adj. p'   = formatC(p.value, format = "e", digits = 2)
  )

knitr::kable(
  onl_ld_sum,
  format    = "latex",
  booktabs  = TRUE,
  caption   = NULL,                    
  align     = c("l","l","r","r","r","r"),
  escape    = TRUE,
  longtable = FALSE,
  linesep   = ""
)

```

## 4. $\Delta$ ONL

```{r}
#| label: delta_onl_interactions
#| include: true
#| echo: false
#| warning: false

data_onl <- readxl::read_excel("Kv82_exp7_all_biomarkers.xlsx", sheet = "onl")

data_onl_delta <- data_onl |>
  dplyr::filter(!is.na(delta)) |>
  dplyr::mutate(
    genotype = factor(genotype, levels = c("Het","KO")),
    side   = factor(side,   levels = c("Inferior","Superior"))
  )

onl_delta_lmm <- lmerTest::lmer(
  delta ~ genotype * side + (1 | mouse),
  data = data_onl_delta
)

# Model: genotype × Side
model_onl_delta <- lmerTest::lmer(
  delta ~ genotype * side + (1 | mouse),
  data = data_onl_delta
)

anova_onl_delta <- anova(model_onl_delta, type = 3)
anova_onl_delta_tidy <- tidy_anova(anova_onl_delta)

tab_iovs(
  anova_onl_delta_tidy,
  caption = "\\textbf{Table S4A. }Delta ONL: Type III ANOVA Results (genotype × side)."
)

```

-   The main effect of genotype on ΔONL thickness was significant (p = 0.013).

-   The main effect of side did not reach statistical significance (p = 0.083), and there was no evidence for a genotype × side interaction (p = 0.58). Therefore, ΔONL data were analyzed with genotype as the primary factor; stratification by side was used for consistent visualization purposes.

```{r}
#| label: tbl_delta_onl_emms
#| include: true
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "\\textbf{Table S4B. }Delta ONL: Estimated marginal means."

data_onl <- readxl::read_excel("Kv82_exp7_all_biomarkers.xlsx", sheet = "onl")

data_delta_onl <- data_onl |>
  dplyr::mutate(
    genotype    = factor(genotype, levels = c("Het","KO")),
    side      = factor(side,   levels = c("Inferior","Superior")),
    delta_onl = light - dark
  ) |>
  dplyr::select(mouse, genotype, side, delta_onl)

delta_onl_lmm <- lme4::lmer(
          delta_onl ~ genotype * side + (1 | mouse), data = data_delta_onl)

delta_onl_emm <- emmeans::emmeans(delta_onl_lmm, ~ genotype | side)


d_onl_emm_df <- as.data.frame(confint(delta_onl_emm, level = 0.95)) |>
  dplyr::transmute(
    Side       = side,
    Genotype     = genotype,
    Mean       = round(emmean, 2),
    SE         = round(SE, 2),
    `lower CI` = round(lower.CL, 2),
    `upper CI` = round(upper.CL, 2)
  ) 


knitr::kable(
  d_onl_emm_df,
  format    = "latex",
  booktabs  = TRUE,
  caption   = NULL,
  align     = c("l","l","r","r","r","r"),
  escape    = TRUE,
  longtable = FALSE,
  linesep   = ""
)

```

```{r}
#| label: tbl_delta_onl_contrasts_genotype
#| include: true
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "\\textbf{Table S4C. }Delta ONL: Het–KO contrasts within side."

# Het–KO contrasts within side (Holm, with 95% CI)
delta_onl_contr_genotype <- emmeans::contrast(
  delta_onl_emm,
  method = "revpairwise",
  by = "side",
  adjust = "holm"
)

delta_onl_sum <- as.data.frame(
  summary(delta_onl_contr_genotype, infer = c(TRUE, TRUE))
) |>
  dplyr::transmute(
    Side        = side,
    `Estimate`  = round(estimate, 2),
    `lower CI`  = round(lower.CL, 2),
    `upper CI`  = round(upper.CL, 2),
    `Holm-adj. p` = formatC(p.value, format = "e", digits = 2)
  )

knitr::kable(
  delta_onl_sum,
  format    = "latex",
  booktabs  = TRUE,
  caption   = NULL,
  align     = c("l","r","r","r","r"),
  escape    = TRUE,
  longtable = FALSE,
  linesep   = ""
)

```
